Did covid encourage assaults in Toronto neighbourhoods in 2020?

### 2022-04-27 outputs/paper/paper.Rmd
commiting final changes and completed assignment

---
title: "Modelling Covid's effect on Assaults in Toronto"
subtitle: "A statistical analysis of how the rise of Covid in 2020 affected assault rates in various downtown neighborhoods in Toronto"
author: 
  "Varun Vijay"
thanks: "Code and data are available at: https://github.com/Varun1005473462/final_folder-main.git."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This paper presents an analysis on the effect of Covid on assault rates in downtown Toronto neighbourhoods in 2020. The paper uses datasets obtained from opendatatoronto to show how Covid had a negative effect on crime rates. The analysis consists of using the statistical programming language R to model the data using regression and constructing various plots and graphs to visually represent the data. THe results obtained will allow us to better predict people's behavior. With the better prediction of people's behaviour, the police can better manage the public order and government will be able to make reasonable and effective policy to reduce the crime rate.\\par\\textbf {Keywords:} Covid, Assaults, Crime,Toronto neighbourhoods, residents, open data toronto"

output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r libraries, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(modelsummary)
library(readr)
library(here)
library(bookdown) # Referencing and report making
library(dplyr) # Data manipulation
library(kableExtra) #Table displaying
library(ggthemes) # More plot themes
library(reshape2) # Dataframe merging
library(janitor) # Cleaning data
library(ggplot2) # To make graphs
library(modelsummary)
knitr::opts_chunk$set(echo = TRUE)
```

# 1 Introduction

After the first reported outbreak of Covid in December 2019, the entire world changed. Covid was quickly declared to be a health emergency of global concern by the World health organization in March of 2020. Governments took measures to restrict the spread of the disease such as imposing lockdowns, travel bans, restrictions on public gatherings, closing down businesses and advising people to isolate at home. Citizens of countries were essentially restricted to their homes and rarely left. Covid had a huge impact on the global economy and had restricted people's actions and movements. However, this paper wants to analyse how Covid has affected one of the most common types of crime, assaults. Assaults are often physical and in a time where physical contact is discouraged, it seems contradictory to assault someone as there is the risk you might catch covid. The question this paper seeks to answer is how the Covid pandemic had affected assault rates in various neighbourhoods in Toronto for the year 2020.

In this paper, we use data on Covid cases and crime statistics in Toronto neighbourhoods for the year 2020 to see how the number of Covid cases in a neighbourhood had affected assualts in that neighbourhood and what factors affected the assault rate the most. We constructed a model using R in order to see how Covid can affect the assault rate in a neighborhood. We also use the dataset to construct various histograms, bar plots and summary statistics about the dataset to improve the quality of our analysis. The results obtained seem to imply that the Covid pandemic had a negative effect on assaults in 2020 as crime rates went down in most downtown neighborhoods. The results obtained contribute to our understanding of how the presence of a common threat that affects everyone such as a pandemic in this case can discourage people from criminal actions. It will allow us to improve our predictions of a person's behavior during such situations. Better predictions of people's behaviors will allow the police to better manage the public order and government to make reasonable and effective policy to reduce the crime rate

The paper first presents an overview of the datasets and the variables we will be using for this study in the data section. The model section presents the regression steps taken to generate the model that shows how Covid affects crime. The results section discusses the various conclusions we can draw from the data and the model and finally the discussion section discusses the strengths and weaknesses of the paper. Any additional details are included in the appendix.

# 2 Data

## 2.1 Data Source and collection

For the dataset, we obtain data using opendatatoronto on crime related statistics, Covid related data pertaining to Toronto neighborhoods.

## 2.2 Data Modification

## 2.3 Data Tables

```{r table 1, echo=FALSE}
options(readr.show_col_types = FALSE)
covid_cases<-
  read_csv(here::here("inputs/data/covid_cases.csv"))
covid_cases<-
  covid_cases |>
  mutate(year=format(covid_cases$`Episode Date`,"%Y"))
covid_cases2020<-
  covid_cases |>
  filter(year == 2020)
covid_cases_gender<-covid_cases2020 %>% group_by(`Neighbourhood Name`, `Client Gender`) %>% count()
covid_cases_gender<-
  covid_cases_gender |>
  pivot_wider(names_from = `Client Gender`,values_fill =0,values_from = n)
covid_cases_outbreak<-covid_cases2020 %>% group_by(`Neighbourhood Name`,`Outbreak Associated`) %>% count()
covid_cases_outbreak<-
  covid_cases_outbreak |>
  pivot_wider(names_from = `Outbreak Associated`,values_fill =0,values_from = n)
covid_cases_final<-merge(covid_cases_gender,covid_cases_outbreak)
covid_cases_final<-
  covid_cases_final |>
  mutate(total_cases=`Outbreak Associated`+Sporadic)
colnames(covid_cases_final)[1]<-"Neighbourhood"
colnames(covid_cases_final)[2]<-"FEMALE_Covid_Cases"
colnames(covid_cases_final)[3]<-"MALE_Covid_Cases"
covid_cases_final<-na.omit(covid_cases_final)
```

```{r table 2, echo=FALSE}
neighbourhood_crime_rates<-
  read_csv(here::here("inputs/data/neighbourhood_crime_rates.csv"))
neighbourhood_crime_rates_assaults<- 
  neighbourhood_crime_rates |>
  select(Neighbourhood,
         Assault_2019,
         Assault_2020)
neighbourhood_crime_rates_assaults<-arrange(neighbourhood_crime_rates_assaults,Neighbourhood)
```

## Obtaining the final dataset.

```{r table 3, echo=FALSE, fig.cap="Final Dataset"}
covid_cases_final<-arrange(covid_cases_final,Neighbourhood)
neighbourhood_crime_rates_assaults<-arrange(neighbourhood_crime_rates_assaults,Neighbourhood)
final_dataset<-
  neighbourhood_crime_rates_assaults |>
  mutate(FEMALE_Covid_CASES=covid_cases_final$FEMALE_Covid_Cases,MALE_Covid_CASES=covid_cases_final$MALE_Covid_Cases,transgender_covid_cases=covid_cases_final$TRANSGENDER,unknown_covid_cases= covid_cases_final$UNKNOWN,other_covid_cases=covid_cases_final$OTHER, Outbreak_Associated=covid_cases_final$`Outbreak Associated`,Sporadic=covid_cases_final$Sporadic,Total_CASES=covid_cases_final$total_cases)
final_dataset
```

## 2.4 Data Visualization

```{r summary 1, echo=FALSE}
final_dataset |>
  datasummary_skim()
```

```{r summary 2,echo=FALSE}
final_dataset |>
  datasummary_correlation()
```

```{r figure 1, echo=FALSE, fig.cap="Assualts in 2020 compared to assaults in 2019"}
colors<-c("Assaults 2019"="red", "Assaults 2020"="blue")
ggplot()+
geom_point(data=final_dataset, mapping=aes(x=Neighbourhood, y=Assault_2019,color="Assaults 2019"),size=1) + 
geom_point(data=final_dataset, aes(x=Neighbourhood, y=Assault_2020,color="Assaults 2020"),size=1)+
  labs(x="Neighbourhoods",
       y="Assaults 2019, Assaults 2020",
       title="Assualts in 2020 compared to assaults in 2019")
```

```{r figure 2, echo=FALSE, fig.cap="Covid cases for each neighborhood in 2020"}
colors<-c("Female"="red", "Male"="blue", "Transgender"="purple","Unknown"="green","Other"="yellow")
ggplot()+
geom_point(data=final_dataset, mapping=aes(x=Neighbourhood, y=FEMALE_Covid_CASES,color="Female"),size=1) + 
geom_point(data=final_dataset, aes(x=Neighbourhood, y=MALE_Covid_CASES,color="Male"),size=1)+
  geom_point(data=final_dataset, aes(x=Neighbourhood, y=transgender_covid_cases,color="Transgender"),size=1)+
  geom_point(data=final_dataset, aes(x=Neighbourhood, y=unknown_covid_cases,color="Unknown"),size=1)+
  geom_point(data=final_dataset, aes(x=Neighbourhood, y=other_covid_cases,color="Other"),size=1)+
  labs(x="Neighbourhoods",
       y="Female, Male, Transgender, Unknown, Other",
       title="Covid cases by gender and neighborhood")
```

@@

$$ y = \beta_0 +\beta_1*x_1+\beta_2*x_2  $$

y= number of assaults in a neighbourhood
x_1 = number of female covid cases in a neighbourhood
x_2 = number of male covid cases in a neighbourhood

Equation \@ref(eq:bayes) seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful.

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

## Table of estimates

```{r model estimates, echo=FALSE}
Terms<-c("Intercept", "Female_covid_cases estimate", "Male_covid_cases estimate", "Number of observation", "R^2", "R^2 adjusted", "AIC", "BIC", "Log.Lik", "F")
values<-c(58.410, -1.174, 2.029, 140, 0.363, 0.354, 1698.6, 1710.3, -845.280, 39.044)
data.frame(Terms, values)
```

$$ y = 58.410	 -1.174*x_1+2.029*x_2  $$

```{r figure , echo=FALSE, fig.cap="Relationship between Covid and Assault rates in 2020"}
final_dataset |>
  ggplot(mapping = aes(x = Total_CASES, y = Assault_2020,color="")) +
  geom_point()
```

```{r, echo=FALSE}
final_dataset |>
  ggplot(aes(x = FEMALE_Covid_CASES+MALE_Covid_CASES, y = Assault_2020)) +
  geom_point() +
  geom_smooth(method = "lm",
              color = "black", 
              linetype = "dashed",
              formula = 'y ~ x') +
  labs(x = "FEMALE_Covid_CASES+MALE_Covid_CASES",
       y = "total_Crime_2020") +
  theme_classic() +
  scale_color_brewer(palette = "Set1")
```

```{r}
plot(final_model)
```

@@

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

@@

## AIC selection
```{r, echo=FALSE}
first_model <-
  lm(Assault_2020 ~FEMALE_Covid_CASES,
     data = final_dataset)
second_model <-
  lm(Assault_2020 ~ MALE_Covid_CASES,
     data = final_dataset)
third_model <-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES,
     data = final_dataset)
fourth_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES+Sporadic,
     data = final_dataset)
fifth_model <-
  lm(Assault_2020 ~ FEMALE_Covid_CASES+MALE_Covid_CASES+Outbreak_Associated,
     data = final_dataset)
sixth_model <-
  lm(Assault_2020 ~ FEMALE_Covid_CASES+MALE_Covid_CASES+Sporadic+Outbreak_Associated,data = final_dataset)
seventh_model<-
   lm(Assault_2020 ~ Sporadic+Outbreak_Associated,data = final_dataset)
eight_model<-lm(Assault_2020 ~ 1,data = final_dataset)
ninth_model<-lm(Assault_2020~Total_CASES,data=final_dataset)
modelsummary(list(first_model,second_model,third_model,fourth_model,fifth_model,sixth_model, seventh_model,eight_model,ninth_model))
```

```{r model, echo=FALSE, message=FALSE, include=FALSE}
final_model<-lm(formula=Assault_2020~FEMALE_Covid_CASES+MALE_Covid_CASES,data=final_dataset)
modelsummary(final_model)
```

# Additional details

>>>>

---
title: "Did covid encourage assaults in Toronto neighbourhoods in 2020?"
subtitle: "A statistical analysis of how the rise of Covid in 2020 affected assault rates in various neighborhoods in Toronto"
author: 
  "Varun Vijay"
thanks: "Code and data are available at: https://github.com/Varun1005473462/final_folder-main.git."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This paper presents an analysis on the effect of Covid on assault rates in Toronto neighbourhoods in 2020. The paper uses datasets obtained from opendatatoronto to show how covid has a positive relation with assaults in a neighbourhood. The analysis consists of using the statistical programming language R to model the data using regression and constructing various plots and graphs to visually represent the data. The results obtained will allow us to better predict people's behavior. With the better prediction of people's behaviour, the police can better manage the public order and government will be able to make reasonable and effective policy to reduce the crime rate.\\par\\textbf {Keywords:} Covid, Assaults, Crime,Toronto neighbourhoods, residents, open data toronto"

output:
  bookdown::pdf_document2
toc: TRUE
bibliography: references.bib
---

```{r libraries, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(modelsummary)
library(readr)
library(here)
library(bookdown) # Referencing and report making
library(dplyr) # Data manipulation
library(kableExtra) #Table displaying
library(ggthemes) # More plot themes
library(reshape2) # Dataframe merging
library(janitor) # Cleaning data
library(ggplot2) # To make graphs
library(modelsummary)
knitr::opts_chunk$set(echo = TRUE, out.width = "65%", out.height = "65%")
```

# Introduction

After the first reported outbreak of Covid in December 2019, the entire world changed. Covid was quickly declared to be a health emergency of global concern by the World health organization in March of 2020. Governments took measures to restrict the spread of the disease such as imposing lockdowns, travel bans, restrictions on public gatherings, closing down businesses and advising people to isolate at home. Citizens of countries were essentially restricted to their homes and rarely left. Covid had a huge impact on the global economy and had restricted people's actions and movements. However, this paper wants to analyse how the rise of Covid has affected one of the most common types of crime, assaults, which are often physical in nature. The question this paper seeks to answer is how the Covid pandemic had affected assault rates in various neighbourhoods in Toronto for the year 2020.

In this paper, we use data on Covid cases and crime statistics in Toronto neighbourhoods for the year 2020 to see how the number of Covid cases in a neighbourhood had affected assaults in 2020 and what factors affected the assault rate the most. We constructed a linear model using R [@citeR] to model covid's effect on assaults. We also used the dataset to construct various plots and summary statistics about the dataset to improve the quality of our analysis. The results obtained seem to imply that the Covid pandemic had a positive effect on assaults in 2020 as there is a positive relation between number of covid cases in a neighbourhood and number of assaults in Toronto.

The results obtained from this paper help contribute to our understanding of how the presence of a common threat that affects everyone such as a pandemic in this case can discourage people from criminal actions. It will allow us to improve our predictions of a person's behavior during such situations. Better predictions of people's behaviors will allow the police to better manage public order and government to make reasonable and effective policy to reduce the crime rate during such situations.

The paper first presents an overview of the datasets and the variables we will be using for this study in the data section. The model section presents the variables chosen for our model and the justification for these variables. The results section discusses the steps taken to generate estimates for our final model and the various conclusions we can draw from  both the data and the model and finally the discussion section discusses the strengths and weaknesses of the paper. Any additional details are included in the appendix. An enhancement is included in appendix C.

# Data

## Data Source and collection

The relevant datasets were obtained using opendatatoronto and show statistics related to Covid cases and to crime figures. The first dataset is about Covid cases in Toronto and records all Covid cases reported since January 2020. The dataset includes all cases related to outbreaks and sporadic cases, those that occur in the community. The data was extracted from the provincial Case & Contact Management System and the variables in the dataset include the neighbourhood the case was reported in, the date it was reported, the client gender, Age group and the classification of the case, whether it was probable or improbable. The full dataset \@ref(table: dataset1) can be found in Appendix B.

The second dataset used for this paper was the dataset related to crime statistics in Toronto neighbourhoods. The dataset contains a list of crime related statistics by neighbourhood and includes figures such as number of assaults, number of car thefts, number of shootings, etc. for the years 2014 to 2020. We are only interested in the number of assaults for 2020. The full dataset \@ref(table: dataset2) can be found in Appendix B.

The datasets were extracted using R [@citeR] and R packages opendatatoronto [@citeopendatatoronto], tidyverse [@tidyverse], dplyr [@dplyr] and knitr [@knitr]. The data was then processed, manipulated and analysed for the purposes of this paper.

## Data Modification

For the purposes of this paper, we are interested only in covid cases that were reported during 2020 since that was when the pandemic first started. So the only cases selected were those reported during 2020. We then focused on the variables Client Gender and the age group of the infected. These variables are the ones that pertain the most to the total number of covid cases in a neighbourhood and so are the ones we will be focusing on for this dataset as they also more important than the others. New variables using the dataset were created such as number of female covid cases, number of male covid cases, number of transgender covid cases, number of other genders and unknown genders covid cases and variables related to the various age groups, less than 19, between 20 to 29, 30 to 39, etc. We also created a new variable where we counted the total number of Covid cases in a neighbourhood. However, we will be excluding all age groups greater than or equal to 50 since it is unlikely that anyone of that age will be assaulting people as they are mostly middle aged or old people. All NA cases were also deleted in order to improve the analysis.

For the crime dataset, we only focused on the neighbourhoods and the number of assaults for the year 2020 since these pertain directly to my paper. We then joined the two datasets together and obtained the final_dataset. All 140 neighbourhoods are included in this dataset and it is the dataset we will be using for the rest of this paper.

The code for the dataset modifications are included in scripts in final-data-prep.R.

## Dataset for this paper

```{r dataset_final, echo=FALSE}
final_dataset<-
  read_csv(here::here("inputs/data/final_dataset.csv"))
final_dataset
```

The final dataset \@ref(table: dataset_final) consists of 140 observations and 12 variables:
Neighbourhood: Name of the neighbourhood,
Assault_2020: number of assaults in that neighbourhood for the year 2020,
FEMALE_Covid_CASES: Number of reported covid cases in that neighbourhood for females,
MALE_Covid_CASES: Number of reported covid cases in that neighbourhood for males,
transgender_Covid_CASES: Number of reported covid cases in that neighbourhood for transgenders,
unknown_covid_cases: Number of reported covid cases in that neighbourhood for people of unknown genders,
other_covid_cases: Number of reported covid cases in that neighbourhood for people of other genders,
<=19: number of reported covid cases for people less than 19 years old,
20 to 29: number of reported covid cases for people between 20 and 29,
30 to 39: number of reported covid cases for people between 30 and 39,
40 to 49: number of reported covid cases for people between 40 and 49,
Total_CASES: total number of covid cases in a neighbourhood.

These 12 variables are the ones that are most pertinent to the topic of this paper. And in this dataset, the number of assaults in a neighbourhood will the response variable for this paper.

## Summary statistics

```{r summary 1, echo=FALSE}

## obtaining summary statistics for the dataset using modelsummary
summary<-final_dataset |>
  datasummary_skim()
summary
```

The above table \@ref(table: summary 1) shows summary statistics for the dataset. As you can see, on average, the number of assaults per neighbourhood in Toronto was 127.5 for 2020 with a standard deviation of 127.5 which means that there was a variation in assaults per neighboruhood. Furthermore, the number of covid cases on average in a neighbourhood is 225.4 with a standard deviation of 206.3 so each neighbourhood has a varying amount of covid cases. So it is possible there is a relation between number of assault cases in a neighbourhood and number of covid cases.

## Correlation statistics for the final dataset

```{r summary 2,echo=FALSE}

## obtaining correlation statistics for the dataset using modelsummary
final_dataset |>
  datasummary_correlation()
```

From the above table \@ref(table: summary 2) we can see that the variables with the strongest correlation with assaults in 2020 are number of female covid cases, number of male covid cases, number of covid cases for age groups 20 to 29, 30 to 39 and 40 to 49. It makes for these variables to relate to number of assaults in a neighbourhood since they are the most important ones since age and gender are incredibly important in such situations when an assault is occurring. We also see that age groups and the number of male and female covid cases are strongly related.

## Assault cases by neighbourhood

```{r figure 1, echo=FALSE,fig.dim = c(8, 6), fig.cap="Assualts in 2020 by neighbourhood"}
## constructing a ggplot for assaults in 2020
ggplot()+
geom_point(data=final_dataset, aes(x=Neighbourhood, y=Assault_2020,color="red"),size=1)+
  labs(x="Neighbourhoods",
       y="Assaults 2020",
       title="Figure 1: Assualts in 2020 by neighbourhood")
```

Fig \@ref(fig: figure 1) shows us the number of assaults in 2020 per neighbourhood. There does seem to be some slight variation in the number of assaults per neighbourhood with some high outliers which match the results shown in the summary table.

## Covid cases by neighbourhood

```{r figure 2, echo=FALSE,fig.dim = c(8, 6), fig.cap="Covid cases for each neighborhood in 2020"}
## constructing a ggplot to see how covid cases vary by neighbourhood
final_dataset |>
  ggplot(mapping=aes(x=Neighbourhood,y=Total_CASES, color="red"))+
  geom_point()+
  labs(x= "Neighbourhoods",
       y="Total number of covid cases in the neighbourhood",
       title="Figure 2: Covid cases for each neighbourhood")
```

As we can see from figure \@ref(fig: figure 2), the number of covid cases per neighbourhood varies and are not similar to each other. Additionally, the variation pattern in the number of covid cases seems to be similar to the pattern in the number of assaults per neighbourhood from figure \@ref(fig: figure 1.) So, the plots support our hypothesis that there is a relation between covid cases and number of assaults in a neighbourhood.

@@

The model we will be using for this paper is:

$$ y = \beta_0 +\beta_1*x_1+\beta_2*x_2 +\beta_3*x_3 +\beta_4*x_4 +\beta_5*x_5 $$

y= number of assaults in a neighbourhood,
x_1 = number of female covid cases in a neighbourhood,
x_2 = number of male covid cases in a neighbourhood,
x_3 = number of covid cases for people between ages 20 and 29
x_4 = number of covid cases for people between ages 30 and 39,
x_5 = number of covid cases for people between ages 40 and 49.

Here, the number of assaults is the response variable and the number of female covid cases, number of male cases, number of covid cases for age group 20 to 29, number of covid cases for age group 30 to 39 and number for 40 to 49 are the explanatory variables.

The reason I have chosen these 5 variables is because from the correlation table \@ref(table: summary 2) in the data section, these variables have some of the strongest correlation with the number of assaults in 2020.

Gender is one of the most important factors in today's world and is an important variable to include. It is always important to take gender into account as it can often be a confounding variable if you do not include it in the model. I have excluded the transgender and unknown variable from the model because those entries were far less than the number of entries for male and female. Since the male and female cases compose the majority of the dataset it is likely that they will have a higher effect on the response variable so we will be using these variables for the model.

People older than 50 would not normally be commiting assaults so we will exclude them from the dataset and also exclude people younger than 20 since they would be teenagers and children. So we will only be focusing on the age groups in the range of 20 to 49 for this model.

@@

Using modelsummary [@modelsummary] and AIC selection process, we were able to obtain the final model and the estimates for the model such as intercept, R^2, estimates for independent variables etc which are included in the table \ref(table: model estimates).

The AIC process and the code for the estimates are located in tables \ref(eq: model). By the process of forwards AIC, we were able to obtain the final model by removing variables from the model to see if it reduced the AIC value. Once we obtained the lowest AIC value possible and could not add more variables, we had obtained my final model.

## Model estimates

```{r model estimates, echo=FALSE}
## creating 2 columns and then merging them to get the dataset
Terms<-c("Intercept", "Female_covid_cases estimate", "Male_covid_cases estimate","20 to 29 estimate", "30 to 39 estimate","40 to 49 estimate", "Number of observation", "R^2", "R^2 adjusted", "AIC", "BIC", "Log.Lik", "F")
values<-c(64.112, -0.918, -0.343,2.06, 4.37,-1.573, 140 ,0.397, 0.375, 1696.9, 1717.5, -841.440, 17.648)
data.frame(Terms, values)
```

## Final model

So, using the estimates obtained, the model we have is:

$$ y = 64.112	-0.918*x_1+-0.343*x_2 +2.060*x_3 +4.370*x_4 -1.573*x_5 $$

The estimates for our model show us that there seems to be a positive relationship between the number of covid cases in a neighbourhood and the number of assaults. When there is an increase in the number of male or female cases in a neighbourhood, then there is a small decrease in the number of assaults. However, if the age group affect was 20 to 29 or 30 to 39, the positive increase in the number of assaults is greater than the decrease due to x1 and x2. Only when the age group is of 40 to 49 does the assaults actually decrease but since two of the age groups have a positive effect on number of assaults, we can say that the number of cases of covid in a neighbourhood will have a positive effect on the number of assaults in that neighbourhood.

So when number of covid cases in a neighbourhood increase, so do the number of assaults. From our analysis of the dataset and the model, this means that in neighbourhoods which had higher covid cases in 2020 also had a higher number of assaults for that year.

## Final model graphs

We will use graphs to confirm our results by modelling covid's effect on assaults in a neighbourhood.

```{r figure 3, echo=FALSE,fig.dim = c(8, 6), fig.cap="Relationship between Covid and Assault rates in 2020"}
final_dataset |>
  ggplot(mapping = aes(x = Total_CASES, y = Assault_2020,color=""))+
  geom_point()+
  labs(x="Total covid cases in a neighbourhood",
       y="Assaults in 2020",
       title="Figure 3: Relation between covid cases and assaults")
```

From figure 3 \@ref(fig:figure 3), we can see that there appears to be a sort of positive relation between the total number of cases in a neighbourhood and the number of assaults in that neighbourhood for the year 2020. Except for a few outliers, it seems that when cases are low, assaults are low and when cases increase, assaults also increase.

```{r figure 4,fig.dim = c(8, 6), fig.cap="Relation between number of assaults in 2020 and the number of female covid cases", echo=FALSE}
final_dataset |>
  ggplot(aes(x = FEMALE_Covid_CASES, y = Assault_2020)) +
  geom_point() +
  geom_smooth(method = "lm",
              color = "black", 
              linetype = "dashed",
              formula = 'y ~ x') +
  labs(x = "FEMALE_Covid_CASES",
       y = "Assaults in 2020",
       title = "Figure 4: Relation between number of assaults in 2020 and the number of female covid cases") +
  theme_classic() +
  scale_color_brewer(palette = "Set1")
```

```{r figure 5, fig.dim = c(8, 6), echo=FALSE}
final_dataset |>
  ggplot(aes(x = MALE_Covid_CASES, y = Assault_2020)) +
  geom_point() +
  geom_smooth(method = "lm",
              color = "black", 
              linetype = "dashed",
              formula = 'y ~ x') +
  labs(x = "MALE_Covid_CASES",
       y = "Assaults in 2020",
       title = "Figure 5: Relation between number of assaults in 2020 and number of male covid cases") +
  theme_classic() +
  scale_color_brewer(palette = "Set1")
```

```{r figure 6,fig.dim = c(8, 6), echo=FALSE}
final_dataset |>
  ggplot(aes(x =`20 to 29`, y = Assault_2020)) +
  geom_point() +
  geom_smooth(method = "lm",
              color = "black", 
              linetype = "dashed",
              formula = 'y ~ x') +
  labs(x = "covid cases in age group 20 to 29",
       y = "Assaults in 2020",
       title = "Figure 6: Relation between number of assaults in 2020 and number of covid cases for age group 20 to 29") +
  theme_classic() +
  scale_color_brewer(palette = "Set1")
```

```{r figure 7,fig.dim = c(8, 6), echo=FALSE}
final_dataset |>
  ggplot(aes(x =`30 to 39`, y = Assault_2020)) +
  geom_point() +
  geom_smooth(method = "lm",
              color = "black", 
              linetype = "dashed",
              formula = 'y ~ x') +
  labs(x = "covid cases in age group 30 to 39",
       y = "Assaults in 2020",
       title = "Figure 7: Relation between number of assaults in 2020 and number of covid cases for age group 30 to 39") +
  theme_classic() +
  scale_color_brewer(palette = "Set1")
```

```{r figure 8,fig.dim = c(8, 6), echo=FALSE}
final_dataset |>
  ggplot(aes(x =`40 to 49`, y = Assault_2020)) +
  geom_point() +
  geom_smooth(method = "lm",
              color = "black", 
              linetype = "dashed",
              formula = 'y ~ x') +
  labs(x = "covid cases in age group 40 to 49",
       y = "Assaults in 2020",
       title = "Figure 8: Relation between number of assaults in 2020 and number of covid cases for age group 40 to 49") +
  theme_classic() +
  scale_color_brewer(palette = "Set1")
```

```{r figure 9,fig.dim = c(8, 6), echo=FALSE}
final_dataset |>
  ggplot(aes(x = FEMALE_Covid_CASES+MALE_Covid_CASES+`20 to 29`+`30 to 39`+`40 to 49`, y = Assault_2020)) +
  geom_point() +
  geom_smooth(method = "lm",
              color = "black", 
              linetype = "dashed",
              formula = 'y ~ x') +
  labs(x = "FEMALE_Covid_CASES+MALE_Covid_CASES+20 to 29+30 to 39+40 to 49",
       y = "Assaults in 2020",
       title = "Figure 9: Relation between number of assaults in 2020 and the final model obtained") +
  theme_classic() +
  scale_color_brewer(palette = "Set1")
```

Figure 4,5,6,7,8 and 9 all show that there seems to be a positive relation between the number of covid cases in a neighbourhood and the number of assaults in that neighbourhood except for a few outliers. When the number of covid cases in a neighbourhood are low so are the number of assaults and when the number of covid cases increase so do the number of assaults increase with it.

```{r figure 10,fig.dim = c(8, 6), fig.cap="Figure 9: Goodness of fit plots", echo=FALSE}
par(mfrow = c(2,2))
model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES+`30 to 39`+`40 to 49`,
     data = final_dataset)
plot(model)
```

We can analyse the goodness of fit for the model from the above 4 plots \ref(fig:figure 10). From the residuals vs fitted plot, we see that linearity for the model does hold and that the independent variables do enter in a linear way. However, from the spread of the residuals we see that variance of our residuals is not constant which means that the model has heteroscedasticity. From the Normal Q-Q plot, we see that our data is normally distributed since the points mostly lie on the dotted line except for a few outliers.

## Conclusion of results

The results obtained from our model and analysis of the dataset seem to imply that the number of covid cases in a neighbourhood had a positive relationship with the number of assaults in that same neighbourhood in the year 2020. The final model we obtained is:

$$ y = 64.112	-0.918*x_1+-0.343*x_2 +2.060*x_3 +4.370*x_4 -1.573*x_5 $$

where 
y= number of assaults in a neighbourhood
x_1 = number of female covid cases in a neighbourhood
x_2 = number of male covid cases in a neighbourhood
x_3= number of covid cases between ages 20 and 29
x_4 = number of covid cases between ages 30 and 39
x_5 = number of covid cases between ages 40 and 49

The number of assaults were higher in neighbourhoods with higher covid cases and were lower in neighbourhoods with lower number of covid cases.

@@

The paper analysed datasets about neighbourhood crime statistics and covid cases reported since January 2020 obtained using opendatatoronto in order to see how number of covid cases in a neighbourhood affect the number of assaults. Here the number of assaults in a neighbourhood was the response variable. The paper focused only on the data during 2020 since that was the year covid was first reported and we were interested in it's initial effects on crime. A linear regression model with several explanatory variables was built using the data obtained and several graphs were created based on this model. Using AIC we were able to build a model and obtain estimates for the various variables.

We found that the variables that affected the response variable were the number of females covid cases in a neighbourhood, number of male covid cases, number of covid cases for age group 20 to 29 and age group 30 to 39. We also found that the number of covid cases in a neighbourhood had a positive relation with the number of assaults in a neighbourhood. Neighbourhoods with higher number of covid cases also had a higher number of assaults. Various graphs plotted using the data confirmed these results for us as they showed a positive and linear relation between number of covid cases in a neighbourhood and number of assaults.

The results obtained supports the theory that pandemics such as covid increase the number of assaults in an area that has been severely afflicted in such situations. Desperation leads to panic which might cause people to inflict violence and assault others. It shows us that when people are placed in situations of high stress, they might panic and resort to violence especially when the stress start to increase and people around them fall ill to some affliction. Our model also showed us that people who were younger and were young adults were more susceptible to panic than older people

The results obtained in this paper could be used to inform police actions in the future at the beginning of situations like Covid where the entire city is impacted. Public order could be maintained by having police increase policing in neighbourhoods that are suffering the most to help protect citizens or have governments make better policy decisions during such situations so that people do not feel stressed and react violently. As an example, they could ensure better care for those in neighbourhoods that are suffering the most from the situation and could focus on younger age groups.

## Weaknesses

There are a few weaknesses in this paper, the datasets used did not include factors such as race of the person with covid which could become a limitation as race might be a confounding factor. However, it might be unethical to try and have race in the model.

Another limitation for the paper is that we were not able to obtain economic data for Toronto neighbourhoods for the year 2020. Economic data such as income of households or families is one of the most important variables to account for when analysing crime since income of a person often dictates whether they would turn to crime or not. Since, we were not able to obtain economic data for the year 2020, it is possible that our paper will be biased in its analysis.

## Next steps

The next steps to improve the model obtained here is to increase the data used for this model. Economic data for Toronto neighbourhoods should be obtained for the year 2020 in order to include these variables in the model. If we are able to obtain such data and include it in the future, we will be able to avoid the effect of confounding variables and will be able to obtain an unbiased model.

Another limitation that could be accounted for is racial data but profiling such data might not be ethical as we will be including race in a model on crime which will lead to racial profiling. This paper recommends that race not be included in such models since it is unethical to model it.

So, in order to fully comprehend people's behaviours in such situation more data is required in order to obtain an unbiased model that will help contribute to out understanding of people's behaviours during high stress situations.

@@

# Appendix A

## AIC selection

For this paper, AIC selection using modelsummary was used to obtain the final model. Several models were constructed and then using modelsummary, the model with the lowest AIC value was selected.

```{r, echo=TRUE}
## the first model constructed using lm with no variables
first_model <-
  lm(Assault_2020 ~1,
     data = final_dataset)
modelsummary(first_model)
```

The AIC obtained from this model is 1757.7, so we will add a variable to see if we can reduce it.

```{r, echo=TRUE}
second_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES,data = final_dataset)
modelsummary(second_model)
```

The AIC obtained from this model is 1745.0 which is less so we can keep the female variable.

```{r, echo=TRUE}
third_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES,data = final_dataset)
modelsummary(third_model)
```

AIC obtained is 1728.4 so we can keep going.

```{r, echo=TRUE}
fourth_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES+transgender_covid_cases,data = final_dataset)
modelsummary(fourth_model)
```

The AIC obtained is 1729.5 so we don't include transgender covid cases

```{r, echo=TRUE}
fifth_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES+`20 to 29`,data = final_dataset)
modelsummary(fifth_model)
```

We keep the 20 to 29 variable since the AIC is lower.

```{r, echo=TRUE}
sixth_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES+`20 to 29`+`30 to 39`,data = final_dataset)
modelsummary(sixth_model)
```

So we move on with this model since the AIC is 1697.2

```{r, echo=TRUE}
seventh_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES+`20 to 29`+`30 to 39`+`40 to 49`,data = final_dataset)
modelsummary(seventh_model)
```

We proceed on with this model and will try to remove the 20 to 29 to see if we get a lower AIC.

```{r, echo=TRUE}
eight_model<-
  lm(Assault_2020 ~FEMALE_Covid_CASES+MALE_Covid_CASES+`30 to 39`+`40 to 49`,data = final_dataset)
modelsummary(eight_model)
```

The AIC is not reduced so we keep 20 to 29.
